\section*{Community benefit (300 words):}

The project involves a new collaboration between Kent (Petricek), Edinburgh
(Cheney, Bach) and Bristol (Wolf), with Turing Fellows from all three partner
institutions contributing, as well as DePaul University in the US (Manik) and
the London Turing office (Perera, Petricek). A wide range of career stages and
disciplines are represented in our investigator team. Our unique selling point
is bringing techniques and perspectives from programming languages to bear on
problems in data visualisation and data science; we see this as a fertile area
for future interdisciplinary research and community-building.

Our work package will bring together Turing researchers interested in
explorability and explainability in data science, and methodological and ethical
issues surrounding data visualisation in particular, such as trustworthiness,
transparency, pedagogy and digital literacy. Our initial efforts will focus on
urban analytics researchers from the Digital Twins SPF programme, but we will
also seek engagement from the Data Science for Science programme, and other SPF
projects dependent on visualisation.

Beyond the Turing network, we envisage our framework being used as part of the
various external projects and collaborations we are involved with. Wolf holds a
fellowship at the Center for Spatial Data Science at University of Chicago and
is co-maintainer of the PySAL spatial analytics library and lead author of
Python libraries for geovisualisation. Bach has collaborated with Glasgow's
Urban Big Data Centre, and historians and archaeologists in the UK, France and
Luxembourg. Cheney has been involved in digital curation and is a research
leader in data provenance. Malik's SciUnit tool is integrated with Hydroshare, a
social platform for sharing code and data serving more than 3500 geoscientists.
Finally, Perera and Petricek are involved in conference and workshop
organisation in live programming, a related (and hot) topic in programming
languages. These will provide a wide variety of opportunities for community
outreach and involvement.
