\section*{Route to Impact (300 words):}

\paragraph{Workshop on explorable data visualisation.} To reach out to
prospective users in the Urban Analytics and Data Science for Science SPF
programmes within the Turing, we will hold a hands-on workshop/tutorial, using
Wrattler, on explorable visualisations. Bach is giving a workshop on data
visualisation at the Turing on 31 May; this will serve as a useful networking
exercise and precursor to our event.

\paragraph{Online gallery of examples/interactive essays.} To generate
excitement beyond the Turing network, we will publish our domain use cases as an
online suite of examples showcasing our approach. We will encourage
contributions from the community, both within and without the Turing. In
collaboration with other relevant TPS projects, we will also aim to publish an
interactive research paper in Google's Distill \cite{goertler19} journal, a
high-profile venue for non-traditional research artefacts, describing the
analysis techniques and design principles underlying our framework.

\paragraph{Standalone self-explaining visualisations.} We will make it possible
to deploy our visualisations as standalone charts or infographics that can be
embedded into web pages or online papers, following the approach taken by
Petricek for The Gamma \cite{petricek17} and leveraging Malik's expertise in
reusable research objects \cite{yuan18}, substantially increasing the potential
reach of our approach beyond practising data scientists to the wider community.

\vspace{8pt}
Naturally, we will also disseminate our research through the usual academic
channels. We will aim for a significant publication at a programming languages,
data visualisation or human-computer interaction conference (perhaps POPL, VIS
or CHI), and one at a data science or quantitative geography/geovisualisation
venue (perhaps SciPy or NACIS). The software will be released under an open
source licence through the Turing's public GitHub repository.
